The Efficient Handling of Guards in the Design of RPython's Tracing JIT

### 2012-08-15 talk/vmil2012/paper.tex
Typos, small fixes, rephrasing

\begin{abstract}
Tracing just-in-time (JIT) compilers record linear control flow paths,
inserting operations called guards at points of possible divergence. These
operations occur frequently generated traces and therefore it is important to
design and implement them carefully to find the right trade-off between
execution speed, deoptimization,
and memory overhead.  In this paper we describe the design decisions about
guards taken in the implementation of the RPython tracing JIT. Furthermore we
measure various properties of guards.
% \o/
\end{abstract}

@@

Our aim is to help understand the design constraints when
implementing guards. Guards have a runtime cost, they take time to execute. On
the other hand, guards are possible deoptimization points, meaning the recorded
and compiled path has to be left returning control to the interpreter. They need
enough associated information to enable rebuilding the interpreter state.
Based on the informal observation that guards are among the most common
operations in the traces produced by RPython's tracing JIT, our
goal is to present concrete numbers for the frequency and the overhead related
to guards, explain how they are implemented in the different levels of this particular
tracing JIT and clarify the rationale behind the design decisions based on the
numbers provided here.

@@

There are several aspects to consider
in the design and optimization of guards, the first aspect is that due to the
large number of guards the memory overhead related to storing the information
needed for deoptimization should be kept low. A second aspect is that
successfully checking guards, i.e. not leaving the compiled trace, – which is
the common case – should be a cheap operation to execute favouring the on-trace
execution speed in contrast to the deoptimization case where the state has to
be rebuilt using the stored information. These constraints and trade-offs are
what make the design and optimization of guards an important and non-trivial
aspect of the low-level design of a tracing just-in-time compiler.

@@

The set of central concepts upon which this work is based is described in
Section~\ref{sec:Background}, such as the PyPy project, the RPython language
and its meta-tracing JIT. Based on these concepts in Section~\ref{sec:Resume
Data} we proceed to describe for RPython's tracing JIT the details of guards in
the frontend related to recording and storing the
information required to rebuild the interpreter state in case of a guard
failure. Once the frontend has traced and optimized a loop it invokes the
backend to compile the operations to machine code, Section \ref{sec:Guards in
the Backend} describes the low-level aspects of how guards are implemented in
the machine specific JIT-backend. The frequency of guards and the overhead associated with the
implementation described in this paper is discussed in
Section~\ref{sec:evaluation}. Section~\ref{sec:Related Work} presents an
overview about how guards are treated in the context of other just-in-time
compilers. Finally Section~\ref{sec:Conclusion} summarizes our conclusions and
gives an outlook on further research topics.

@@

The RPython language and the PyPy project were started in 2002 with the goal of
creating a Python interpreter written in a high level language, allowing easy
language experimentation and extension. PyPy is now a fully compatible
alternative implementation of the Python language, that is on average about 5 times faster than the reference implementation. The
implementation takes advantage of the language features provided by RPython
such as the provided tracing just-in-time compiler described below.

RPython, the language and the toolset originally developed to implement the
Python interpreter have developed into a general environment for experimenting
and developing fast and maintainable dynamic language implementations. There
are, besides the Python interpreter, implementations of Prolog, Javascript, R,
Smalltalk among other that are written in RPython at different levels of
completeness.

>>>>

\begin{abstract}
Tracing just-in-time (JIT) compilers record linear control flow paths,
inserting operations called guards at points of possible divergence. These
operations occur frequently in generated traces and therefore it is important to
design and implement them carefully to find the right trade-off between
execution speed, deoptimization,
and memory overhead.  In this paper we describe the design decisions about
guards taken in the implementation of the RPython tracing JIT. Furthermore we
measure various properties of guards.
% \o/
\end{abstract}

@@

Our aim is to help understand the constraints when
implementing guards. Guards have a runtime cost, they take time to execute. On
the other hand, guards are possible deoptimization points, meaning the recorded
and compiled path has to be left returning control to the interpreter. They need
enough associated information to enable rebuilding the interpreter state.
Based on the informal observation that guards are among the most common
operations in the traces produced by RPython's tracing JIT, our
goal is to present concrete numbers for the frequency and the overhead related
to guards, explain how they are implemented in the different levels of this particular
tracing JIT and clarify the rationale behind the design decisions based on the
numbers provided here.

@@

There are several aspects to be taken into account w.r.t. guards, the first aspect is that due to the
large number of guards the memory overhead related to storing the information
needed for deoptimization should be kept low. A second aspect is that
successfully checking guards, i.e. not leaving the compiled trace, – which is
the common case – should be a cheap operation to execute favouring the on-trace
execution speed in contrast to the deoptimization case where the state has to
be rebuilt using the stored information. These constraints and trade-offs are
what make the design and optimization of guards an important and non-trivial
aspect of the construction of a tracing just-in-time compiler.

@@

The set of central concepts upon which this work is based are described in
Section~\ref{sec:Background}, such as the PyPy project, the RPython language
and its meta-tracing JIT. Based on these concepts in Section~\ref{sec:Resume
Data} we proceed to describe for RPython's tracing JIT the details of guards in
the frontend. In this context the frontend is concerned with recording and storing the
information required to rebuild the interpreter state in case of a guard
failure. Once the frontend has traced and optimized a loop it invokes the
backend to compile the operations to machine code, Section \ref{sec:Guards in
the Backend} describes the low-level aspects of how guards are implemented in
the machine specific JIT-backend. The frequency of guards and the overhead associated with the
implementation described in this paper is discussed in
Section~\ref{sec:evaluation}. Section~\ref{sec:Related Work} presents an
overview about how guards are treated in the context of other just-in-time
compilers. Finally Section~\ref{sec:Conclusion} summarizes our conclusions and
gives an outlook on further research topics.

@@

The RPython language and the PyPy project were started in 2002 with the goal of
creating a Python interpreter written in a high level language, allowing easy
language experimentation and extension.\footnote{\url{http://pypy.org}} PyPy is now a fully compatible
alternative implementation of the Python language, that is on average about 5 times faster than the reference implementation. The
implementation takes advantage of RPython's language features
such as the provided tracing just-in-time compiler described below.

RPython, the language and the toolset originally created to implement the
Python interpreter have developed into a general environment for experimenting
and developing fast and maintainable dynamic language implementations. Besides
the Python interpreter there are several language implementation at different
levels of completeness, e.g. for Smalltalk, Prolog, JavaScript and R.
